---
title: "Assignment 4 ETC5513"
author: "XXXX"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 2
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
bibliography: references.bib 
biblio-style: plainnat
subtitle: Team name
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center} 
  \includegraphics[width=5in,height=6in]{Images/sam-albury-oA7MMRxTVzo-unsplash.jpg}\LARGE\\}
- \posttitle{\end{center}}
- \usepackage{fontawesome}
- \usepackage[most]{tcolorbox}
- \usepackage{xcolor}
- \usepackage{sectsty}
- \sectionfont{\color{blue}}
- \usepackage{verbatim}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning= FALSE)
```

```{r libraries}
library(tidyverse)
library(knitr)
library(readr)
library(lubridate)
library(bookdown)
library(usmap)
library(kableExtra)
library(gridExtra)
theme_set(theme_minimal())
```

```{r cleaning-data}
# read data
storm17 <- read_csv("data/StormEvents_details-ftp_v1.0_d2017_c20200121.csv")
storm18 <- read_csv("data/StormEvents_details-ftp_v1.0_d2018_c20200317.csv")
storm19 <- read_csv("data/StormEvents_details-ftp_v1.0_d2019_c20200516.csv")

# change date format
clean <- function(x) {
x$BEGIN_DATE_TIME <- dmy_hms(x$BEGIN_DATE_TIME) 
x$END_DATE_TIME <- dmy_hms(x$END_DATE_TIME)
x$BEGIN_DAY <- mday(x$BEGIN_DATE_TIME)
x$END_DAY <- mday(x$END_DATE_TIME)
x$BEGIN_TIME <- format(as.POSIXct(x$BEGIN_DATE_TIME) ,format = "%H:%M:%S") 
x$END_TIME <- format(as.POSIXct(x$END_DATE_TIME) ,format = "%H:%M:%S") 
x$MONTH_NAME <- month(x$BEGIN_DATE_TIME, label=TRUE)

#unselect column with NA value; unselect BEGIN MONTH and END MONTH because there are already MONTH column
x <- x %>% select(-BEGIN_YEARMONTH, -END_YEARMONTH, -EPISODE_ID, -WFO, -CZ_TIMEZONE, -SOURCE, -MAGNITUDE_TYPE,-CATEGORY, -FLOOD_CAUSE,-TOR_OTHER_WFO,- TOR_OTHER_CZ_STATE, -TOR_OTHER_CZ_FIPS, - TOR_OTHER_CZ_NAME, - BEGIN_AZIMUTH, -BEGIN_LOCATION, -END_AZIMUTH, -END_LOCATION, -EPISODE_NARRATIVE, -EVENT_NARRATIVE, -DATA_SOURCE) 
}

storm17 <- clean(storm17)
storm18 <- clean(storm18)
storm19 <- clean(storm19)
storm_all <- rbind(storm17,storm18,storm19)
```

\clearpage

# Introduction
Storm can disturb the surface of earth by thunderstorm, snowstorm, rainstorm, sandstorm, flood and so on to cause the road impassibility, the damage of properties and even the injuries and death. Current research shows that hurricanes in the eastern North Pacific can provide water to parts of the southwestern United States and parts of Mexico. More than half of Japan’s rainfall comes from typhoons, but at the same time strong convective weather such as hail can also damage buildings and cars parked on the road. , And seriously affected the output of crops such as corn and wheat. This report will focus on the storms happened in America from 2017 to 2019 to analyse it from locations, damages and frequency.
\clearpage

# Data

## Data source

We download the dataset from Storm Events Database data set which contains 31 variables and 187139 observations in total from NOAA <https://www.ncdc.noaa.gov/stormevents/>.

## Data limitation

One of the defects of this dataset is that there is no further division of the scale of disasters, because the losses, casualties and even distribution areas caused by disasters of different scales are different, for example, it is quite difficult to distinguish between small versus large hail through this dataset. So, if there is a further division of the scale of disasters, it will be more conducive to analysis.

\clearpage

# Methodology

\clearpage

# Section 1
From the Table \@ref(tab:stormtb) the top 10 numbers of storm events in America are all concentrate in 2017 and 2019 in the past three years. The thunderstorm wind happened most frequently, which is 18617 cases in 2019 and 16472 cases in 2017, after that is hail followed by 10398 in 2017 and 9013 in 2019.
```{r stormtb, echo=FALSE}
event19 <- storm19 %>%
  select(EVENT_TYPE, YEAR) %>%
  group_by(EVENT_TYPE, YEAR) %>%  
  summarise(count=n())

event18 <- storm18 %>%
  select(EVENT_TYPE, YEAR) %>%
  group_by(EVENT_TYPE, YEAR)

event17 <- storm17 %>%
  select(EVENT_TYPE, YEAR) %>%
  group_by(EVENT_TYPE, YEAR) %>%  
  summarise(count=n())

events_3years <- merge(merge(event19, event18, all=TRUE), event17, all=TRUE)%>%
  arrange(desc(count))%>%
  head(10)

knitr::kable(events_3years, caption = "Top10 storm events in 3 years ")%>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

Table \@ref(tab:countstb) shows the top 10 numbers of storm events happened  in America in 2019. first of all is thunderstorm wind happened 18617 times, after that is hail with 9013 cases and flood with 4949 cases and flash flood followed by 4072 cases. The other six are winter weather, high wind, winter storm, heavy snow, marine thunderstorm wind and tornado in order.
```{r countstb, echo=FALSE}
event_counts <- storm19 %>%
  select(STATE, EVENT_TYPE, BEGIN_LAT, BEGIN_LON) %>%
  group_by(EVENT_TYPE) %>%
  summarise(count=n())%>%
  arrange(desc(count))%>%
  head(5)

knitr::kable(event_counts, caption = "Top5 storm event frequency in 2019")%>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

Figure  \@ref(fig:locationmap) shows the distributions of the most frequently happened five storm events' locations in 2019. It is clear to see that theses storm events happened more in eastern of America and some in the middle of America. 
```{r locationmap, fig.cap = "Top5 storm event locations in 2019", fig.path = "Images/", echo=FALSE, fig.align = "center"}
library(ggmap)
library(ggthemes)
library(ggplot2)
usa_bbox <- c(-130,
              20,
              -60,
              50)
usa_map <- get_map(location = usa_bbox, source = "osm")
storm_locations <- data.frame(storm19 %>%
  select(STATE, EVENT_TYPE, BEGIN_LAT, BEGIN_LON) %>%
  filter(EVENT_TYPE %in%c("Thunderstorm Wind", 
                    "Hail",
                    "Flood",
                    "Flash Flood",
                    "Winter Weather"),
         BEGIN_LON != "NA",
         BEGIN_LAT != "NA"))
ggmap(usa_map) +
  geom_point(data= storm_locations, aes(x= BEGIN_LON, y = BEGIN_LAT, color = EVENT_TYPE),
              alpha = 0.3, size = 0.5) +
  theme_map()
```

Figure \@ref(fig:TWmonth) shows the thunderstorm wind locations in America in 2019 of different months. It is clear to see that the most active months of thunderstorm wind were from May to August. Furthermore, the locations of theses events happened from middle and east of America slowly transformed to the east from July to October. 
```{r TWmonth, fig.cap = "Thunderstorm locations by month in 2019", fig.path = "Images/", echo=FALSE, fig.align = "center"}
Thunderstorm_locations <- data.frame(storm19 %>%
  select(STATE, EVENT_TYPE, MONTH_NAME, BEGIN_LAT, BEGIN_LON) %>%
  filter(EVENT_TYPE %in%c("Thunderstorm Wind"),
         BEGIN_LON != "NA",
         BEGIN_LAT != "NA"))
         
ggmap(usa_map) +
  geom_point(data= Thunderstorm_locations, aes(x= BEGIN_LON, y = BEGIN_LAT),
             color = "blue",alpha = 0.1, size = 0.3) +
  facet_wrap(~ MONTH_NAME)
```

Table \@ref(tab:statestb) displayed the number of thunderstorm wind happed in different states of America in 2019. As we can see from the table, except for Kansas, Texas and Missouri, the top ten states are all located in the east of America.
```{r statestb, echo=FALSE}
thunderstorm_counts <- storm19 %>%
    filter(EVENT_TYPE %in%c("Thunderstorm Wind")) %>%
  select(STATE, EVENT_TYPE, DEATHS_DIRECT, BEGIN_LAT, BEGIN_LON) %>%
  group_by(STATE) %>%
  summarise(count=n())%>%
  arrange(desc(count))%>%
  head(10)

knitr::kable(thunderstorm_counts, caption = "Top10 states by number of thunderstorm in 2019")%>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

Figure \@ref(fig:monthplot) shows that the number of  thunderstorm wind occurred in different states by month. The thunderstorm wind most frequently happened from May to August. However the peak of the number of such cases were different from some states, August is the peak month of some eastern America states like New York, Pennsylvania and Virginia, for the middle of America states like Texas and Missouri, the peak months were May and June.
```{r monthplot, fig.cap = "The nuber of thunderstorm events of top10 states in diffenrent month", fig.path = "Images/", echo=FALSE, fig.align = "center"}
thunderstorm_time <- storm19 %>%
    filter(EVENT_TYPE %in%c("Thunderstorm Wind")) %>%
  select(STATE, EVENT_TYPE, MONTH_NAME) %>%
  filter(EVENT_TYPE %in%c("Thunderstorm Wind"),
         STATE %in%c("PENNSYLVANIA",
                     "VIRGINIA",
                     "TEXAS",
                     "NEW YORK",
                     "OHIO",
                     "KANSAS",
                     "NORTH CAROLINA",
                     "GEORGIA",
                     "MISSOURI",
                     "SOUTH CAROLINA")) %>%
  group_by(MONTH_NAME, STATE) %>%
  summarise(count=n())%>%
ggplot(aes(x = MONTH_NAME, y = count)) +
  geom_point() +
  facet_wrap(~ STATE)
thunderstorm_time
```



\clearpage

# Section 2


\clearpage

# Section 3

## The trend of occurence of the top five most frequent events changing by year 

In this section, the main purpose is to explore the trend of occurence of the top five most frequent events changing by year both across the whole USA and in each state. First, create a table and group the data by "STATE" and "YEAR". Then count the times of occurence for all events and rename the name of this column as "FREQ" to present the frequency. And next, create a new table to arrange the data by descending order with displaying the top 5 events that happen most frequently. Then, use "right_join" to connect these two tables. Finally, create two figures to display the trend of occurence of the top 5 event changing by year.
```{r frequencytrend, echo = FALSE}
storm_frequency_state <- storm_all %>%
  group_by(STATE, YEAR) %>%
  count(EVENT_TYPE) %>%
  rename(FREQ = n) 
```

```{r tablefrequency, echo = FALSE}
kable(storm_frequency_state[1:10,],
      booktabs = TRUE, 
      caption = "The frequency of the events") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = F,
                latex_options = c("hold_position"))
```


```{r frequencytrendnew, echo = FALSE}
storm_frequency_state_new <- storm_all %>%
  group_by(EVENT_TYPE) %>%
  count(EVENT_TYPE) %>%
  rename(FREQ = n) %>%
  arrange(desc(FREQ))%>%
  head(5) %>%
  select(EVENT_TYPE)
```

```{r tablefrequencynew, echo = FALSE}
kable(storm_frequency_state_new[1:5,],
      booktabs = TRUE, 
      caption = "Top five most frequent events") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = F,
                latex_options = c("hold_position"))
```

```{r topnewtable, echo = FALSE}
top5_newtable <- right_join(storm_frequency_state, storm_frequency_state_new, by = "EVENT_TYPE") %>%
  group_by(STATE)
```

```{r rightjointable, echo = FALSE}
kable(top5_newtable[1:10,],
      booktabs = TRUE, 
      caption = "Occurence of top five most frequent events") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = F,
                latex_options = c("hold_position"))
```

```{r  frequencyplot,fig.height = 30, fig.width= 30, echo = FALSE, fig.cap="Occurence of top five most frequent events changing by year"}
ggplot(top5_newtable,
       aes(x= YEAR, 
           y = FREQ, group = EVENT_TYPE, color = EVENT_TYPE)) +
  geom_line() +
  scale_y_log10() +
  theme(
  legend.title = element_text(color = "blue", size = 60),
  legend.text = element_text(color = "green", size = 60),
  axis.title.x = element_text(size = 40),
  axis.title.y = element_text(size = 40),
  axis.text.x = element_text(size = 30),
  axis.text.y = element_text(size = 30)
  )
```


Table \@ref(tab:tablefrequency) displays the first ten lines of the frequency of all events. And Table \@ref(tab:tablefrequencynew) displays the top five most frequent events. And Table \@ref(tab:rightjointable) displays the first ten lines of the final results of the occurence of top five most frequent events. It can be seen that top 5 events have occurred in some regions in three years for example, thunderstorm wind has occurred in Alabama in three yearsfrom 495 times to 630 times , while it has only occurred twice in Alaska in 2019.
Figure \@ref(fig:frequencyplot) presents the occurence of top five most frequent events across the whole USA, we could see that thunderstorm wind showed an increasing trend during three years. And hail showed a slow growth in 2017-2018, followed by a sharp decline in 2018-2019. As for the winter weather, it has been on a downward trend for three years, but the slowdown from 2017 to 2018 was larger than that from 2018 to 2019. What's more, flash flood kept growing, but slowed down between 2018 and 2019. At last, flood kept a slow decline at first, then turned into a significant rise. 
We could see that most events have a similar trend, but hail has decreased in 2018 to 2019 years, which is possibly due to increased melting level heights and greater atmospheric instability.[@dessensberthetsanchez2015] Correspondingly, the number of floods increased from 2018 to 2019, which is increasingly common due to years of relative sea level increases and El Nino.[@floodingdaysin2018]

\clearpage

# Section 4

Based on the calculation of the number of tornadoes that occur each year, the United States becomes a country with more than 1000 events each year. @USTornad80 Table \@ref(tab:table-tornado) shows the number of tornadoes from 2017 to 2019 in the United States.
The table made by `kable` function from `knitr` package. @knitr 

```{r table-tornado}
storm_all %>% filter(EVENT_TYPE %in% c("Tornado")) %>%
  group_by(YEAR, EVENT_TYPE) %>%
  tally() %>%
  kable(caption = "Total number of Tornado that occurs from 2017-2019 in United States") %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

We use `ggplot2` package to map the distribution overtime. @ggplot2 To see the number of tornadoes that occur each month during these three years can be seen in Figure \@ref(total-f-scale). From this plot, it can be seen that most tornadoes occur on average in the range of April and May. 2019 recorded a high number of tornadoes in May compared to the previous two years.

```{r total-f-scale, fig.cap= "Tornado frequency from 2017-2019 in USA", fig.width =7, fig.height=4}
# Total F scale
storm_all$YEAR<- as.character(storm_all$YEAR)
storm_all %>% filter(EVENT_TYPE %in% c("Tornado")) %>%
  group_by(YEAR, MONTH_NAME) %>%
  tally() %>%
  ggplot(aes(x=MONTH_NAME,
             y= n,
             group = YEAR,
             colour=YEAR))+
  geom_line() +
  geom_point() +
  theme(legend.position = "bottom")
```

\clearpage

Figure \@ref(fig:map-tornado) shows the map of the United States with the highest number of tornadoes during 2017-2019. The map plotted using `plot_usmap` function from `usmap` package. @usmap 

```{r map-tornado, fig.cap= "United States map of total tornado in each state", fig.height=3, fig.width=6}
statepop1 <- storm_all %>% filter(EVENT_TYPE %in% c("Tornado")) %>%
  group_by(STATE_FIPS, STATE) %>%
  tally() %>%
  summarise(total = sum(n))

# make sure fips table is char and has the name fips to match the usmap function
statepop1$STATE_FIPS <- as.character(statepop1$STATE_FIPS)
statepop1 <- statepop1 %>% rename(fips=STATE_FIPS)

plot_usmap(data = statepop1, values = "total", label=TRUE, label_color = "Black")+
  scale_fill_continuous(
    low = "white", high = "red", name = "Total Tornado 2017-2019", label = scales::comma
  ) + theme(legend.position = "right")
```

Each tornado that occurs has a scale. Based on dataset documentation, enhanced Fujita Scale describes the strength of the tornado based on the amount and type of damage caused by the tornado. @documentation
The F scale listed as Table \@ref(tab:scale) :

```{r scale}
scale_f <- data_frame("Scale" = c("EF0","EF1","EF2","EF3", "EF4", "EF5", "EFU"),
                      "Category" = c("Light","Moderate","Significant","Severe", "Devastating","Incredible","Unknown"),
                      "Speed" = c("40 – 72 mph","73 – 112 mph","113 – 157 mph","158 – 206 mph","207 – 260 mph","261 – 318 mph","Unknown"))
kable(scale_f, caption = "Level of Tornadoes") %>%
    kableExtra::kable_styling(latex_options = "hold_position")
```
When there is no tornado damage, but there is evidence a tornado existed (citation)

In Table \@ref(tab:tornado-state), Texas is the state that has been hit the most by Tornadoes—followed by Kansas, Alabama, Missouri, and Minnesota. The top four states have experienced Tornadoes on an EF4 scale from 2017 to 2019. The table style use `kableExtra` package. @kableExtra

```{r tornado-state}
storm_all %>% filter(EVENT_TYPE %in% c("Tornado")) %>%
  group_by(STATE, TOR_F_SCALE) %>%
  tally() %>%
  pivot_wider(names_from = TOR_F_SCALE, values_from = n) %>%
  arrange(desc(EF0)) %>%
  head(5) %>%
  kable(caption = "Tornado frequency by scale for each state in USA from 2017-2019") %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

As the state with the highest number of tornadoes, we analyze Texas deeper. Figure \@ref(fig:tornado-texas) shows the number of tornadoes that occur overtime based on scale. In general, the number of tornadoes varies each year. EF0 occurs almost every year, but in 2019, the number of EF0 occurs more than 40 times in May. In 2018, the number of tornadoes on the EF1 and EF2 scales was less than in 2017 and 2019. The EF4 happened once in 2018.

```{r tornado-texas, fig.cap= "Tornado scale per month in Texas", fig.width = 7, fig.height=3}
storm_all %>% filter(EVENT_TYPE == "Tornado", STATE == "TEXAS") %>%
  group_by(TOR_F_SCALE, MONTH_NAME, YEAR, STATE) %>%
  tally() %>%
  ggplot(aes(x=MONTH_NAME,
             y = n, 
             group= YEAR,
             colour = YEAR))+
  geom_line()+
  geom_point()+
  facet_wrap(~TOR_F_SCALE) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom") +
  labs(x = "Month", y= "Total")
```

Figure \@ref(fig:tornado-texas-2019) tells what time of day a tornado with a particular scale occurs in Texas in 2019. Generally, there are differences in each month. May is the most prominent month with almost every day some tornadoes occur.

```{r tornado-texas-2019, fig.cap= "Tornado by hour distribution that happened in Texas in 2019", fig.width = 7, fig.height=3.1}
storm_all %>% filter(EVENT_TYPE == "Tornado", STATE == "TEXAS", YEAR == "2019") %>%
  mutate(Time = hour(BEGIN_DATE_TIME)) %>%
  group_by(MONTH_NAME, Time, TOR_WIDTH, TOR_F_SCALE) %>%
  ggplot(aes(x=wday(BEGIN_DATE_TIME, label=TRUE),
             y=Time,
             colour=TOR_F_SCALE)) +
  geom_point() +
  facet_wrap(~MONTH_NAME, ncol = 4,
             labeller = labeller(MONTH_NAME = label_wrap_gen(20))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.3),
        legend.position = "bottom") +
  labs(x = "Days", y = "Time")
```

\clearpage

# Conclusion
The most active months of thunderstorm wind were from May to August. Furthermore, the locations of theses events happened from middle and east of America slowly transformed to the east from July to October.
\clearpage



