---
title: "Assignment 4 ETC5513"
author: "Yan Ma"
output:
  pdf_document:
    citation_package: natbib
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 2
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
bibliography: references.bib 
biblio-style: dinat
subtitle: Team name
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center} 
  \includegraphics[width=5in,height=6in]{Images/sam-albury-oA7MMRxTVzo-unsplash.jpg}\LARGE\\}
- \posttitle{\end{center}}
- \usepackage{fontawesome}
- \usepackage[most]{tcolorbox}
- \usepackage{xcolor}
- \usepackage{sectsty}
- \sectionfont{\color{blue}}
- \usepackage{verbatim}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning= FALSE)
```

```{r libraries}
library(tidyverse)
library(knitr)
library(readr)
library(lubridate)
library(bookdown)
```

```{r cleaning-data}
# read data
storm17 <- read_csv("data/StormEvents_details-ftp_v1.0_d2017_c20200121.csv")
storm18 <- read_csv("data/StormEvents_details-ftp_v1.0_d2018_c20200317.csv")
storm19 <- read_csv("data/StormEvents_details-ftp_v1.0_d2019_c20200516.csv")

# change date format
clean <- function(x) {
x$BEGIN_DATE_TIME <- dmy_hms(x$BEGIN_DATE_TIME) 
x$END_DATE_TIME <- dmy_hms(x$END_DATE_TIME)
x$BEGIN_DAY <- mday(x$BEGIN_DATE_TIME)
x$END_DAY <- mday(x$END_DATE_TIME)
x$BEGIN_TIME <- format(as.POSIXct(x$BEGIN_DATE_TIME) ,format = "%H:%M:%S") 
x$END_TIME <- format(as.POSIXct(x$END_DATE_TIME) ,format = "%H:%M:%S") 
x$MONTH_NAME <- month(x$BEGIN_DATE_TIME, label=TRUE)

#unselect column with NA value; unselect BEGIN MONTH and END MONTH because there are already MONTH column
x <- x %>% select(-BEGIN_YEARMONTH, -END_YEARMONTH, -EPISODE_ID, -WFO, -CZ_TIMEZONE, -SOURCE, -MAGNITUDE_TYPE,-CATEGORY, -FLOOD_CAUSE,-TOR_OTHER_WFO,- TOR_OTHER_CZ_STATE, -TOR_OTHER_CZ_FIPS, - TOR_OTHER_CZ_NAME, - BEGIN_AZIMUTH, -BEGIN_LOCATION, -END_AZIMUTH, -END_LOCATION, -EPISODE_NARRATIVE, -EVENT_NARRATIVE, -DATA_SOURCE) 
}

storm17 <- clean(storm17)
storm18 <- clean(storm18)
storm19 <- clean(storm19)
storm_all <- rbind(storm17,storm18,storm19)
```

\clearpage

# Introduction

\clearpage

# Methodology

\clearpage

# Section 1


\clearpage

# Section 2


\clearpage

# Section 3


\clearpage

# Section 4: Injuries and Deaths

## Direct injuries and deaths by month and year

```{r fig1, fig.cap = "Total direct injuries by month"}
storm_all %>% 
  group_by(YEAR, MONTH_NAME) %>% 
  mutate(injury_tot = sum(INJURIES_DIRECT)) %>% 
  ggplot(aes(x = MONTH_NAME, y = injury_tot, fill = MONTH_NAME)) + 
  geom_col(position = "dodge") + 
  coord_flip() + 
  facet_grid(~ YEAR) + 
  guides(fill=FALSE) + 
  ggtitle("Total direct injuries by month")
```

```{r fig2, fig.cap = "Total direct deaths by month"}
storm_all %>% 
  group_by(YEAR, MONTH_NAME) %>% 
  mutate(death_tot = sum(DEATHS_DIRECT)) %>% 
  ggplot(aes(x = MONTH_NAME, y = death_tot, fill = MONTH_NAME)) + 
  geom_col(position = "dodge") + 
  coord_flip() + 
  facet_grid(~ YEAR) + 
  guides(fill=FALSE) + 
  ggtitle("Total direct deaths by month")
```

## Direct injuries and deaths by state

```{r fig3, fig.cap = "Total direct injuries by state", fig.height=8}
storm_all %>% 
  group_by(YEAR, STATE) %>% 
  mutate(injury_tot_state = sum(INJURIES_DIRECT)) %>% 
  ggplot(aes(x = STATE, y = injury_tot_state, fill = STATE)) + 
  geom_col(position = "dodge") + 
  coord_flip() + 
  facet_grid(~ YEAR) + 
  guides(fill=FALSE) + 
  ggtitle("Total direct injuries by state") + 
  theme(axis.text.x = element_text(size=12), 
        axis.text.y = element_text(size=2))
```

```{r fig4, fig.cap = "Total direct deaths by state", fig.height=8}
storm_all %>% 
  group_by(YEAR, STATE) %>% 
  mutate(death_tot_state = sum(DEATHS_DIRECT)) %>% 
  ggplot(aes(x = STATE, y = death_tot_state, fill = STATE)) + 
  geom_col(position = "dodge") + 
  coord_flip() + 
  facet_grid(~ YEAR) + 
  guides(fill=FALSE) + 
  ggtitle("Total direct deaths by state") + 
  theme(axis.text.x = element_text(size=12), 
        axis.text.y = element_text(size=2))
```

## Direct injuries and deaths by major event type

```{r tab1, tab.cap = "Major event types cause direct injuries over years"}
library(kableExtra)
storm_all %>% 
  group_by(EVENT_TYPE) %>% 
  summarise(injury_event = sum(INJURIES_DIRECT)) %>% 
  arrange(desc(injury_event)) %>% 
  head(5) %>% 
  kable(caption = "Major event types cause direct injuries over years",  booktabs = T) %>% 
  kable_styling(latex_options = c("striped", "hold_position"))

major_types_injury <- c("Tornado", "Heat", "Thunderstorm Wind", "Excessive Heat", "Lightning")
```

```{r fig5, fig.cap = "Total direct injuries by event type"}
storm_all %>% 
  filter(EVENT_TYPE == major_types_injury) %>% 
  group_by(YEAR, EVENT_TYPE) %>% 
  mutate(injury_tot_type = sum(INJURIES_DIRECT)) %>% 
  ggplot(aes(x = EVENT_TYPE, y = injury_tot_type, fill = EVENT_TYPE)) + 
  geom_col(position = "dodge") + 
  facet_wrap(~ YEAR) + 
  guides(fill=FALSE) + 
  ggtitle("Total direct injuries by event type") + 
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

```{r tab2, tab.cap = "Major event types cause deaths over years"}
library(kableExtra)
storm_all %>% 
  group_by(EVENT_TYPE) %>% 
  summarise(death_event = sum(DEATHS_DIRECT)) %>% 
  arrange(desc(death_event)) %>% 
  head(5) %>% 
  kable(caption = "Major event types cause deaths over years",  booktabs = T) %>% 
  kable_styling(latex_options = c("striped", "hold_position"))

major_types_death <- c("Excessive Heat", "Flash Flood", "Heat", "Rip Current", "Wildfire")
```

```{r fig6, fig.cap = "Total direct deaths by event type"}
storm_all %>% 
  filter(EVENT_TYPE == major_types_death) %>% 
  group_by(YEAR, EVENT_TYPE) %>% 
  mutate(death_tot_type = sum(DEATHS_DIRECT)) %>% 
  ggplot(aes(x = EVENT_TYPE, y = death_tot_type, fill = EVENT_TYPE)) + 
  geom_col(position = "dodge") + 
  facet_wrap(~ YEAR) + 
  guides(fill=FALSE) + 
  ggtitle("Total direct injuries by event type") + 
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

\clearpage

# Conclusion

\clearpage

# Citation examples

[More styles for natbib here](https://www.overleaf.com/learn/latex/Natbib_bibliography_styles)


 @BC64 and this is another article about COVID @bai2020presumed and I can also cite R packages as follows @ggplot2
